\mode*
\subsection{Conditionals}

A common control structure taught in CS1 is if- and else-statements, where 
the students learn how to create easy conditionals that controls the 
progress of the programme. How a programme will understand and execute an if
-statement is something that students have misconceptions about. A severe 
misconception that \textcite{Plass2015Variables} found was that some 
students believe that an if-statement can control if the programme will 
keep on executing or shut down, depending on if the statement is true or 
false. Students believe that an if-statement that is false will terminate 
the programme, even though a quit-statement has not been introduced. 
Another misconception is that when writing an if and else statement, both 
if and else will be executed, even when the if-statement is true \autocite{
MisconceptionsSurvey2017}.

Both misconceptions concern the same object of learning: \emph{how control flows
through an if--else statement}. The critical aspect that the students have not
discerned is that the condition merely \emph{selects} which branch runs --- at
most one branch executes, and the program then continues with whatever follows
the statement. A false condition neither terminates the program nor causes both
branches to run. We can make this aspect discernible with variation theory.

\begin{description}
  \item[Contrast] We keep one program invariant and vary only the truth value of
    the condition. The same statement after the conditional runs in both cases
    (so a false \mintinline{python}{if} does not terminate the program), and
    exactly one of the two branches runs (so the \mintinline{python}{else} does
    not also run when the \mintinline{python}{if} is true). Only the
    highlighted value varies.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1}]{python}
        temperature = 30
        if temperature > 20:
            print("It is warm.")
        else:
            print("It is cold.")
        print("Take care!")
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1}]{python}
        temperature = 10
        if temperature > 20:
            print("It is warm.")
        else:
            print("It is cold.")
        print("Take care!")
      \end{minted}
    \end{minipage}
    \newline
    Tracing the output makes the invariant visible: the left program prints
    \enquote{It is warm.} and the right prints \enquote{It is cold.}, but
    \emph{both} print \enquote{Take care!} afterwards.

  \item[Generalisation] We keep the critical value invariant --- exactly one
    branch runs and execution continues --- while varying the non-critical
    shape of the conditional. Adding an \mintinline{python}{elif} branch, or
    nesting, does not change the rule.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}{python}
        if temperature > 20:
            print("It is warm.")
        elif temperature > 0:
            print("It is chilly.")
        else:
            print("It is freezing.")
        print("Take care!")
      \end{minted}
    \end{minipage}

  \item[Fusion] Finally we fuse the two aspects --- which branch runs \emph{and}
    that execution continues --- by letting each branch assign to the same
    variable and using it after the conditional. The student must attend to the
    selection and the continuation simultaneously to predict the output.

    \hfill
    \begin{minipage}[t]{0.6\columnwidth}
      \begin{minted}{python}
        if temperature > 20:
            advice = "shorts"
        else:
            advice = "a jacket"
        print(f"Wear {advice}.")
      \end{minted}
    \end{minipage}
\end{description}

The same critical aspect --- that exactly one branch runs --- also underlies a
related antipattern. When two cases are mutually exclusive, students sometimes
write them as two separate \mintinline{python}{if} statements rather than an
\mintinline{python}{if}--\mintinline{python}{else}. \textcite{Bosse2021Antipatterns}
catalogue this as \emph{not using \enquote{else} where it would be appropriate}:
\textcquote{Bosse2021Antipatterns}{The structure can give the expected result, but
since only one of the selections can be true, the second would only need to be
checked if the first were false}. Expressing the exclusivity with an
\mintinline{python}{else} both states that intent and avoids re-testing.

A further confusion concerns the operator inside the condition itself.
\textcite{Bosse2021Antipatterns} found students writing \mintinline{python}{=}
where \mintinline{python}{==} is meant, since
\textcquote{Bosse2021Antipatterns}{using only an equals sign \enquote{=} is
attribution and not a comparison}. This is the mirror image of the misconception
\textcite{Plass2015Variables} reports: there an assignment is \emph{read} as a
test of equality, whereas here a test of equality is \emph{written} as an
assignment. The object of learning is the distinction between the two operators
--- \mintinline{python}{=} binds a name to a value, while \mintinline{python}{==}
asks whether two values are equal --- which in Python surfaces at once, as
\mintinline{python}{if x = 5:} is a syntax error.

A syntax error that is common is when students try to chain conditions in
an if-statement, for example \mintinline{python}{if x != a or b}, where the 
correct statement should be \mintinline{python}{if x != a or x != b} 
\autocite{GuoMarkelZhang2020}. This misconception is believed to originate 
from the way the statement is read out loud as \enquote{if x is not equal to a 
or b}, which in mathematical terms is the right way to state it. 

Here the object of learning is \emph{how a boolean operator combines
conditions}. The critical aspect is that each operand of
\mintinline{python}{or} (or \mintinline{python}{and}) must be a
\emph{complete} comparison: \mintinline{python}{x != a or x != b} is two
comparisons joined by \mintinline{python}{or}, whereas
\mintinline{python}{x != a or b} joins the comparison \mintinline{python}{x != a}
with the bare value \mintinline{python}{b}, whose truthiness is then tested. To
make this discernible we keep the values invariant and vary only whether the
right-hand side is a full comparison.

\begin{description}
  \item[Contrast] The same values \mintinline{python}{x, a, b}, varying only the
    right operand of \mintinline{python}{or}. With all three equal, the left
    program prints nothing --- correctly, since \mintinline{python}{x} differs
    from neither --- while the right program still prints, because the truthy
    \mintinline{python}{b} makes its condition true regardless of
    \mintinline{python}{x}. The misconception version thus reports a difference
    that is not there, revealing that the comparison did not distribute.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={2}]{python}
        x, a, b = 5, 5, 5
        if x != a or x != b:
            print("x differs from a or b")
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={2}]{python}
        x, a, b = 5, 5, 5
        if x != a or b:
            print("x differs from a or b")
      \end{minted}
    \end{minipage}

  \item[Generalisation] Keeping the rule invariant --- each operand is a complete
    comparison --- we vary the comparison operator and the operands, so the
    student sees the rule is not specific to \mintinline{python}{!=}.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}{python}
        if score == 1 or score == 2:
            print("low")
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}{python}
        if 0 < x and x < 10:
            print("in range")
      \end{minted}
    \end{minipage}
\end{description}
