\mode*
\section{Method}

This study has two parts. The first is a \emph{literature review} that maps the
misconceptions novices hold in introductory programming
(\cref{sec:method-litreview}). The second is an \emph{analysis} of those
misconceptions through the lens of variation theory
(\cref{sec:method-vt-analysis}). The misconceptions surfaced by the review are
the phenomenographic observations on which the variation-theory analysis builds.

\subsection{Literature review}
\label{sec:method-litreview}

To map misconceptions in introductory programming we conducted the review in two
phases. The first phase was a manual, semi-structured review carried out by hand
in a single database. The second phase re-ran the review systematically with a
tool-supported, multi-database search; we then compared the two so as to
evaluate what a tool-assisted, partly automated search adds over a careful
manual one. The full, reproducible protocol for the second phase is given in
\cref{app:litprotocol}.

\subsubsection{Phase one: manual semi-structured review}

A semi-structured literature review is used when the studied area is broad and
has been conceptualised in many different ways \autocite{Snyder2019}. When
conducting a semi-structured literature review the purpose will be to find 
significant findings in the area, however it is not mandatory, nor perhaps 
possible, to review all articles relevant to the researched subject. 
Nevertheless, it is still important to keep the review transparent by stating 
which search words and criterias that have been used when browsing the articles 
available on the subject \autocite{Snyder2019}.
For this literature review of misconceptions the following criterias have 
been used:
\begin{itemize}
\item The word \emph{misconceptions} should be in the title of the article.
\item The article should be focused on \emph{Python} or other languages that 
have similarities to Python. Articles focused on object-oriented languages 
in 
general might also be relevant.
\item The focus in the article should be on introductory programming for 
high 
school or higher education.
\item Articles with the purpose of deciding which programming language is 
the 
most efficient one to use when teaching introductory programming have been 
sifted out because of the irrelevance for the purpose of this study.
\item Articles which revolve around a tool that measures students' knowledge 
instead of common misconceptions have also been sifted out.
\item An iteration with one search word has been stopped when ten articles 
have had the same findings or if ten articles in a row have been irrelevant.
\end{itemize}
\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
Database & Search words\\
\midrule
\multirow{3}{4em}{Google scholar} & common misconceptions in intro to 
programming \\
& common misconceptions cs1 \\
& programming misconceptions student mistakes \\
\bottomrule
\end{tabular}
\caption{Databases and search words used in review}
\label{databasesandwords}
\end{table}
In this first phase the search was carried out by hand in a single database,
Google Scholar. The search words used are presented in \cref{databasesandwords}.
The misconceptions found in each chosen article are presented in
\cref{misconceptions}. This manual phase yielded the set of papers that form our
initial misconception catalogue, and it serves as the \emph{baseline} against
which we compare the systematic phase below.

\subsubsection{Phase two: systematic tool-supported review}

To make the review more systematic, transparent, and reproducible, we re-ran it
with the \texttt{scholar} literature-search tool \autocite{scholar}. Rather than
a single database queried by hand, \texttt{scholar} issues each query to several
bibliographic databases at once, records every query verbatim, and supports
large-language-model (LLM) assisted screening against an explicit set of
inclusion and exclusion criteria. We searched OpenAlex, arXiv, Web of Science,
Scopus, and IEEE~Xplore (Semantic Scholar and DBLP were configured but did not
contribute to the retained set in this run; see \cref{app:litprotocol}). The
refined criteria relax the manual phase's requirement that \emph{misconceptions}
appear in the title to title \emph{or} abstract, while keeping the focus on
concrete misconceptions, errors, and difficulties at the introductory level in
Python or a language whose constructs transfer to Python.

The search retrieved \num{117} unique papers; LLM-assisted screening against the
criteria kept \num{24} and discarded \num{93}, the latter most often as
\emph{not focused on misconceptions} or as \emph{intervention} and
\emph{AI-tool} studies that do not characterise misconceptions. Every retained
reference carries a provenance record (how it was found and the verbatim passage
that supports the misconception it is cited for); the full query set, screening
counts, and provenance method are documented in \cref{app:litprotocol}.

\subsubsection{Comparing the two phases}
\label{sec:method-comparison}

Treating the manual phase's papers as a baseline, we compared what each method
retrieved. The two proved markedly \emph{complementary} rather than redundant:
of the manual baseline's DOI-identifiable misconception papers, the systematic
search recovered only the single most-cited survey
\autocite{MisconceptionsSurvey2017}, while contributing a largely disjoint set
of additional relevant papers that the manual review had not surfaced. Much of
this divergence is explained by database coverage: the manual review found
specific, often older CS-conference papers (a strength of the
computer-science-specific indexes that were unproductive in our systematic run),
whereas the systematic search over general-purpose indexes surfaced a broader
and more recent set. We read this not as one method dominating the other but as
evidence that a tool-assisted, multi-database search \emph{extends} a manual
review and should be paired with citation snowballing (deferred here) to recover
the construct-specific literature. \Cref{app:litprotocol} reports the comparison
in full.

\subsection{Analysis with variation theory}
\label{sec:method-vt-analysis}

Each misconception found in earlier research will be analysed through
variation theory. The analysis has the purpose of understanding how one can 
teach the different aspects revolving around the misconception at hand, in 
order to avoid this misconception in the future. 
