\mode*
\subsection{Problem solving}

\subsubsection{Role in the syllabus}

Problem solving is, in a sense, the apex skill of CS1: taking a problem
description, decomposing it into sub-problems, implementing each, and assembling
the pieces into a working program. It is also where many students fall short.
The ITiCSE working group led by \textcite{Lister2004Tracing} established that, at
the end of an introductory course, a large fraction of students still cannot
program. A natural explanation is a missing problem-solving ability; but the
working group's evidence points one level lower, to the more basic skills that
problem solving rests on:
\blockcquote{Lister2004Tracing}{many students do not know how to program at the
conclusion of their introductory courses. A popular explanation for this
incapacity is lack of the ability to problem-solve [\ldots] An alternative
explanation is that many students have a fragile grasp of both basic programming
principles and the ability to systematically carry out routine programming
tasks, such as tracing [\ldots] suggesting that these skills are a prerequisite
for problem-solving.}

This matters for our analysis: if tracing and a firm grasp of individual
constructs are prerequisites for problem solving, then teaching against the
construct-level misconceptions catalogued above is also groundwork for problem
solving.

\subsubsection{Difficulties that can occur}

Two difficulties recur. The first is mathematical content.
\textcite{KumarVeerasamy2016} found that students did markedly worse on
math-related tasks than on others:
\blockcquote{KumarVeerasamy2016}{This study analysis also explored that novices
of programming struggled in writing code for math-related Questions 6 and 7.
Nearly 66\% of students did not do well in the mathematical problem-based
questions though explained and allowed to surf the Internet to seek for more
details during the exam hours. A neo-Piagetian theory of cognitive development
stated that students who are at the concrete operational stage struggle to write
large programs with partial specifications, although they can write small
programs from well-defined specifications.}

The second is tracing and execution order. Students struggle to follow code
step by step and to reason about the order in which a program will execute
\parencite[p.~20]{Sleeman1984}, a difficulty also documented by
\textcite{Kurvinen2016}. Both feed directly into the problem-solving gap that
\textcite{Lister2004Tracing} identified, since a student who cannot trace a short
program reliably has little hope of assembling a larger one.
