\mode*
\subsection{Repetitions}

In CS1 students usually learn about repetitions, which includes for- and 
while-loops and in some cases recursion.

Loop constructions can be hard to trace and understand for novice students, 
for instance when a loop starts, ends and what is repeated and not repeated 
in the loop \autocite{Sekiya2013,KumarVeerasamy2016,Kaczmarczyk2010}. This 
was something \textcite{Sleeman1984} also realised when studying high 
school students writing and debugging loop-structures. A common 
misconception that the students had was that if the loop contained a print-
statement, the students thought that the only thing repeated inside the 
loop was the string they saw in the terminal. The difficulties students 
have in tracing the code linearly when entering a loop is according to
\textcite{KumarVeerasamy2016} because of the lack of understanding the
students have of the looping technique and the amount of cognitive skills
the tracing takes. \textcite{Eckert2022Loops} similarly found that loops
generally, and nested loops in particular, pose difficulties for novices, and
identified loop misconceptions tied to how the loop is represented.

The object of learning here is \emph{what happens on each iteration of a loop}.
The critical aspect hidden in the misconception above is that the loop body is
re-executed on each pass with the loop variable \emph{rebound} to a new element
--- so what repeats is the body, not merely the fixed string that happened to be
printed. We make this discernible by keeping the iterable invariant and varying
only whether the body refers to the loop variable.

\begin{description}
  \item[Contrast] The same loop over the same list, varying only whether the
    body refers to the loop variable. The left program prints the same string
    three times --- consistent with the misconception that a fixed string is
    what repeats --- while the right one prints \mintinline{python}{10},
    \mintinline{python}{20}, \mintinline{python}{30}, showing that the body is
    re-executed each pass with the loop variable rebound.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={2}]{python}
        for x in [10, 20, 30]:
            print("Hello")
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={2}]{python}
        for x in [10, 20, 30]:
            print(x)
      \end{minted}
    \end{minipage}

  \item[Generalisation] Keeping the critical value invariant --- the body
    repeats and the loop variable takes each element in turn --- we vary the
    \emph{kind} of iterable (a list, a \mintinline{python}{range}, a string).
    \citeauthor{Eckert2022Loops}'s finding that misconceptions are tied to the
    loop's representation motivates exactly this: vary the surface form while
    holding the meaning constant \autocite{Eckert2022Loops}.

    \hfill
    \begin{minipage}[t]{0.3\columnwidth}
      \begin{minted}{python}
        for x in [1, 2, 3]:
            print(x)
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.3\columnwidth}
      \begin{minted}{python}
        for x in range(1, 4):
            print(x)
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.3\columnwidth}
      \begin{minted}{python}
        for c in "abc":
            print(c)
      \end{minted}
    \end{minipage}
\end{description}

When students define and use loops, \textcite{GuoMarkelZhang2020} found
three common misconceptions that the students had about the loop-statement, 

\begin{enumerate}
    \item A misconception about which variable in a loop defined as 
      \mintinline{python}{for item in items}, that is supposed to be use to 
      extract different information.
    \item The misconception that \mintinline{python}{for i in 100} will iterate 
      a hundred times, even though Python require a specified range.
    \item When defining a while-loop, the misconception is that one can write
      \mintinline{python}{while i <= 100} without initialising
      \mintinline{python}{i} beforehand, and that \mintinline{python}{i} will
      automatically increase with 1 inside the loop.
\end{enumerate}

The second and third of these share an object of learning: \emph{what a loop
needs in order to run}. For a \mintinline{python}{for} loop the critical aspect
is that the value after \mintinline{python}{in} must be an \emph{iterable}; for a
\mintinline{python}{while} loop it is that the loop variable must be both
initialised before the loop and updated inside it. A contrast that varies only
the critical aspect makes each requirement visible, because the misconception
version fails observably.

\begin{description}
  \item[Contrast] For the \mintinline{python}{for} loop we vary only what follows
    \mintinline{python}{in}: an integer (which is not iterable, raising
    \mintinline{python}{TypeError: 'int' object is not iterable}) versus a
    \mintinline{python}{range}.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1}]{python}
        for i in 100:
            print(i)
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1}]{python}
        for i in range(100):
            print(i)
      \end{minted}
    \end{minipage}
    \newline
    For the \mintinline{python}{while} loop we vary only whether the loop
    variable is initialised and updated. Without them the loop never starts
    (a \mintinline{python}{NameError}) or never ends; with them it counts as
    intended.

    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1}]{python}
        while i <= 100:
            print(i)
      \end{minted}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.45\columnwidth}
      \begin{minted}[highlightlines={1,3}]{python}
        i = 0
        while i <= 100:
            i = i + 1
      \end{minted}
    \end{minipage}
\end{description}

Another difficult part of the loop technique is to understand how an if-
statement inside a loop is executed. \textcite{Sekiya2013} found in their 
studies that the combination of the two control structures created 
misconceptions. For instance the students thought that the variables in the 
conditional part of the loop-construction was control variables or the 
output from the loop. The students in the studies often got confused and 
started to misplace the different variables that are defined when writing 
an if-statement in a for-loop.

When a conditional is placed inside a loop, two critical aspects must be
discerned \emph{simultaneously}: the loop variable that advances each iteration,
and any variable used by the conditional to accumulate or decide. This is a
fusion situation --- the misconception arises precisely because the student
attends to one role and mistakes it for the other.

\begin{description}
  \item[Fusion] We keep the earlier counting loop and fuse it with a conditional
    that maintains a separate variable. The student must track both the loop
    variable \mintinline{python}{x} (which ranges over the data) and the
    accumulator \mintinline{python}{total} (which the conditional updates) at the
    same time to predict the result.

    \hfill
    \begin{minipage}[t]{0.55\columnwidth}
      \begin{minted}[highlightlines={2,4}]{python}
        total = 0
        for x in [3, 8, 1, 9, 4]:
            if x > 5:
                total = total + x
        print(total)
      \end{minted}
    \end{minipage}
\end{description}
